---
title: "Week 6 Assignment - genetic Algorithm"
author: "Vaishnavi Tapetla"
date: "2025-03-02"
output: html_document
---

```{r}
library(data.table)
library(geosphere)
library(GA)
library(purrr)
library(sf)

## Load the population data
pop.data <- data.table::fread("C:/Users/Others/Downloads/Mo_pop_Sim/Mo_pop_Sim.csv")
str(pop.data)
frac <- 0.0001
small.data <- pop.data[sample(1:nrow(pop.data),
                               size = round(nrow(pop.data) * frac),
                               replace = F), ## extract a sample of randomlychosen 1% rows
                        ]  ## and choose all columns

## Load the FQHC data
data_path <- 'MO_2018_Federally_Qualified_Health_Center_Locations'
fqhc.data <- data.table(as.data.frame(st_read("C:/Users/Others/Downloads/MO_2018_Federally_Qualified_Health_Center_Locations/MO_2018_Federally_Qualified_Health_Center_Locations",                     'MO_2018_Federally_Qualified_Health_Center_Locations')))
```

1. One approach can be to use the average closeness of a location from all of the residences - those locations that are the closest to all of the residences, on average, are the ones that are more easily accessible, and therefore would/should be selected. What is the assumption that is implicit here?

Approach - Calulating Average Distance
```{r}

fqhc_coords <- data.frame(lon = fqhc.data$Longitude, lat = fqhc.data$Latitude)
# Extract coordinates from population sample
pop_coords <- data.frame(lon = small.data$long, lat = small.data$lat)

# Function to calculate average distance from a FQHC to all population points
calculate_avg_distance <- function(fqhc_index, pop_coords, fqhc_coords) {
  fqhc_point <- fqhc_coords[fqhc_index, ]
  distances <- distGeo(p1 = fqhc_point, p2 = pop_coords)
  return(mean(distances))
}
```




2. Another approach is to consider locations that have the highest densities of residences within a 30-mile radius (e.g., if location 1 has 59 residences and location 2 has 84 residences in a 30-mile radius, location 1 is to be considered "better" - why? because it can serve more people)

Approach - Calculating Population Density
```{r}
# Function to calculate population density within a radius
calculate_population_density <- function(fqhc_index, pop_coords, fqhc_coords, radius = 30 * 1609.34) { # 30 miles in meters
  fqhc_point <- fqhc_coords[fqhc_index, ]
  distances <- distGeo(p1 = fqhc_point, p2 = pop_coords)
  return(sum(distances <= radius))
}
```


```{r}
#Calculating fitness for each FQHC using average distance
avg_distances <- sapply(1:nrow(fqhc_coords), calculate_avg_distance, pop_coords = pop_coords, fqhc_coords = fqhc_coords)

#Calculating fitness for each FQHC using population density
population_densities <- sapply(1:nrow(fqhc_coords), calculate_population_density, pop_coords = pop_coords, fqhc_coords = fqhc_coords)

# Normalize fitness scores 
normalized_avg_distances <- 1 / avg_distances # Lower distance is better, so take inverse
normalized_densities <- population_densities / max(population_densities)

# Combine fitness scores
combined_fitness <- normalized_avg_distances + normalized_densities

# Find the 8 FQHC locations with highest fitness
top_8_indices <- order(combined_fitness, decreasing = TRUE)[1:8]

# Print the top 8 FQHCs
top_8_fqhcs <- fqhc.data[top_8_indices,]
print(top_8_fqhcs)
```

```{r}
#best average distance locations
top_8_avg_distance_indices <- order(avg_distances, decreasing = FALSE)[1:8]
top_8_avg_distance_fqhcs <- fqhc.data[top_8_avg_distance_indices,]
print(top_8_avg_distance_fqhcs)
```

```{r}
#8 best density locations
top_8_density_indices <- order(population_densities, decreasing = TRUE)[1:8]
top_8_density_fqhcs <- fqhc.data[top_8_density_indices,]
print(top_8_density_fqhcs)
```

```{r}
# Define the fitness function
fitness_function <- function(indices) {
  indices <- as.integer(indices)
  selected_fqhcs <- fqhc_coords[indices, ]

  # Calculate the combined fitness for the selected FQHCs
  total_fitness <- sum(combined_fitness[indices])
  
  return(total_fitness)
}

# Run Genetic Algorithm
ga_result <- ga(
  type = "permutation",
  fitness = function(x) fitness_function(x), 
  lower = 1, 
  upper = nrow(fqhc_coords),
  popSize = 50, 
  maxiter = 50,
  run = 10,
  pmutation = 0.2,
  elitism = 5,
  parallel = FALSE
)

# Get the best solution
best_solution_indices <- ga_result@solution[1,]
best_fqhcs <- fqhc.data[best_solution_indices, ]

# Print selected FQHC locations
print(best_fqhcs)
```




Population density spanning a 30-mile radius serves as the main factor in determining how fit each FQHC location is for serving patients. My selection bases on the premise that expanding patient reach is the main objective for accessing specialized mental health services. Our target of higher population density areas ensures greater access to the total population density. The method includes an assumption that mental health service requirements are equal among all population members and that densely populated areas need more support. The model requires that a 30-mile distance serves as sufficient for patients to access mental health care providers. Average distance is an accessible measure for tracking universal service exposure yet lacks specific information about actual patient numbers who benefit from treatment services. Density optimization selects maximum coverage and widespread influence above average distance reduction within the area under consideration.

The fitness function for this project would determine the residential population count present in a 30-mile radius surrounding each FQHC facility. I would measure the geodesic distance between each FQHC location coordinates and the coordinates of all residential locations in the sampled population data. After converting each FQHC distance to meters I would determine the number of people who live within 30 miles from the medical facility. The specified radius reveals the population density through the population count determination. Maximization of this fitness measure occurs during evaluation. An increased number of residents indicates better accessibility for mental health care since it shows numerous people living near service areas. Our objective is to optimize the FQHC placement by finding locations that minimize resident travel distances to reach the facilities which serve the largest population numbers thereby making resources stretch further and benefiting the community most effectively.


